---
title: "REBOOT: A Brief Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REBOOT: A Brief Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r}
library(Reboot)
```

To reduce vignette build time, code chunks are displayed but not evaluated during package compilation.

## Overview

REBOOT is an R package for identifying and evaluating prognostic molecular signatures (genes/transcripts)
using bootstrap resampling and penalized LASSO Cox regression.

The package is organized into three main workflows:

- Module I: molecular signature discovery (`rebootRegression()`)
- Module II: survival analysis (`rebootSurvival()`)
- Complete workflow (`rebootComplete()`)

This vignette provides a brief step-by-step guide to running the REBOOT pipeline and exploring its outputs.

---

## Input data

### Expression and survival data

The main input should contain (in this order):

- Survival status (`OS`)
- Survival time (`OS.time`)
- Expression values (genes or transcripts)

Each row represents a sample, and each column represents a molecular feature.

A toy dataset bundled with the package is used below:

```{r}
example_file <- system.file("extdata", "toy_expression.tsv", package = "Reboot")

toy_data <- read_reboot_table(example_file)

head(toy_data[, 1:5])
```

Alternatively, a built-in dataset can be loaded directly:

```{r}
data(toy_expression)

head(toy_expression[, 1:5])
```

Example structure (sample IDs as rownames):

```text
             OS   OS.time   Transcript1   Transcript2   Transcript3
Sample1      1    1200      5.2           3.1           0.0
Sample2      0    980       2.4           1.8           0.5
```

---

### Molecular signature

The `signature` argument used in `rebootSurvival()` and `rebootComplete()` can be provided as:

- A `reboot_signature` object generated by `rebootRegression()`
- A user-defined data.frame
- An external tab-separated file

The signature should contain (in this order):

- The features, either genes or transcripts (`feature`)
- Their respective coefficients from a regression survival analysis (`coefficient`)

Example structure:

```text
feature        coefficient
ENST00001      0.231
ENST00002     -0.412
ENST00003      0.118
```

---

### Clinical data for multivariate analysis

When `multivariate = TRUE` in `rebootSurvival()` or `rebootComplete()`, an additional clinical table must be provided using the `clinin` argument.

The clinical table should contain:

- Sample identifiers matching the expression table
- Binary clinical covariates suitable for multivariate Cox regression

A toy dataset bundled with the package is used below:

```{r}
clin_file <- system.file("extdata", "toy_clinics.tsv", package = "Reboot")

clin_data <- read_reboot_table(clin_file)

head(clin_data)
```

Alternatively, a built-in dataset can be loaded directly:

```{r}
data(toy_clinics)

head(toy_clinics)
```

Example structure (sample IDs as rownames):

```text
             Age    Stage   Gender
Sample1      Old    High    Male
Sample2      Young  Low     Female
```

---

## Module I - Molecular signature discovery

The `rebootRegression()` function identifies robust molecular signatures using filtering steps, feature selection, and bootstrap LASSO Cox regression.

```{r}
set.seed(123)

reg_result <- rebootRegression(
  filein = example_file,
  outprefix = "my_analysis",
  bootstrap = 100,
  groupsize = 10,
  percentagefilter = 0.3,
  variancefilter = 0.01,
  followup = NULL,
  type = "transcript",
  force = TRUE,
  plots = TRUE,
  table = TRUE,
  saveJSON = TRUE,
  saveRDS = TRUE,
  report = TRUE,
  log = TRUE
)
```

For reproducible analyses, users are encouraged to define a random seed before running REBOOT workflows.

When `plots = TRUE` and/or `table = TRUE`, REBOOT automatically exports plots and result tables to the working directory. If these options are disabled (default),
users can manually generate specific outputs using the visualization and export helper functions, as specified at the end of each workflow section.

Moreover, user are able to export metadata information in JSON format, such as parameters used, when `saveJSON = TRUE`. If this option is disabled (default),
users can manually generate the file using the specific helper function, specified at the end of this tutorial.

Similarly, users can export a summary report (in PDF or HTML) of the entire REBOOT pipeline when `report = TRUE`. If this option is disabled (default),
users can manually produce the report using the specific helper function, specified at the end of this tutorial.

Finally, users are provided with the flexibility of saving the results of a specific analysis in an RDS file when `saveRDS = TRUE`. If this option is disabled (default),
users can manually save (and load) the results generated by REBOOT using the specific helper functions specified at the end of this tutorial.

---

## Inspecting regression results

The result is an S3 object of class `reboot_signature`.

```{r}
summary(reg_result)
print(reg_result)
as.data.frame(reg_result)
coef(reg_result)
plot(reg_result)
```

---

## Regression visualization utilities

Plots can also be generated manually:

```{r}
reboot_histogram(reg_result$signature, "my_histogram")

reboot_lollipop(reg_result$signature, "my_lollipop")
```

---

## Module II - Survival analysis

The `rebootSurvival()` function applies a molecular signature to survival data and performs univariate and optional multivariate survival analyses.

```{r}
surv_result <- rebootSurvival(
  filein = example_file,
  signature = reg_result,
  outprefix = "my_survival",
  multivariate = TRUE,
  clinin = clin_file,
  roc = TRUE,
  variancefilter = 0.01,
  followup = NULL,
  p.cutoff = 0.2,
  bootstrap = 100,
  force = TRUE,
  plots = TRUE,
  table = TRUE,
  saveJSON = TRUE,
  saveRDS = TRUE,
  report = TRUE,
  log = TRUE
)
```

---

## Inspecting survival results

The result is an S3 object of class `reboot_survival`.

```{r}
summary(surv_result)
print(surv_result)
as.data.frame(surv_result)
coef(surv_result)
plot(surv_result)
```

---

## Survival visualization utilities

Additional survival-related plots can also be generated manually. Below are a few examples:

```{r}
reboot_km(data = surv_result$expression_data, cutoff = surv_result$score_cutoff, pval = surv_result$survival$univariate$table$log.rank.pvalue, outprefix = "my_km")
```

If ROC was used to define the score cutoff:

```{r}
reboot_roc(x = surv_result$roc_result, outprefix = "my_roc")
```

If multivariate analysis is performed:

```{r}
reboot_forest(model_object = surv_result$survival$multivariate$model, outprefix = "my_forest")
```

---

## Complete workflow

The `rebootComplete()` function sequentially runs both REBOOT modules:

1. Molecular signature discovery
2. Survival analysis

The molecular signature generated during the regression step is automatically passed to the survival module.

```{r}
complete_result <- rebootComplete(
  filein = example_file,
  outprefix = "my_complete_analysis",
  bootstrap = 100,
  groupsize = 10,
  percentagefilter = 0.3,
  variancefilter = 0.01,
  followup = NULL,
  type = "transcript",
  multivariate = TRUE,
  clinin = clin_file,
  roc = TRUE,
  p.cutoff = 0.2,
  force = TRUE,
  plots = TRUE,
  table = TRUE,
  saveJSON = TRUE,
  saveRDS = TRUE,
  report = TRUE,
  log = TRUE
)
```

The parameters `bootstrap`, `variancefilter`, and `followup` are applied to both modules,
although they may operate differently during the regression and survival analysis steps.

---

## Inspecting complete workflow results

The result is an S3 object of class `reboot_complete`.

```{r}
summary(complete_result)
print(complete_result)
coef(complete_result)
plot(complete_result)
```

Module-specific results can also be accessed directly:

```{r}
complete_result$regression

complete_result$survival
```

---

## Import utilities

REBOOT provides helper functions for importing tabular data:

```{r}
my_example_file <- read_reboot_table("path/to/example_file.tsv")
my_sig <- read_reboot_table("path/to/example_signature.txt", row.names = FALSE)
```

---

## Export utilities

The package also provides helper functions for exporting results.

### Molecular signatures

```{r}
write_reboot_signature(reg_result, "my_signature.txt")
```

### Calculated scores

```{r}
write_reboot_score(surv_result, "my_scores.tsv", clinics = TRUE)
```

### Univariate survival results

```{r}
write_reboot_univariate(surv_result, "my_uniCox.txt")
```

### Multivariate survival results

```{r}
write_reboot_multivariate(surv_result, "my_multiCox.txt")
```

### JSON with parameters and additional execution metadata

```{r}
write_reboot_metadata(module = "rebootComplete", outprefix = "my_json.json", command = "none", parameters = list(...))
```

### PDF/HTML REBOOT report
```{r}
write_reboot_report(object = complete_result, format = "PDF", author = "Dr. Your Name", file = "my_report.pdf")
```

---

## Returned object structure

### Molecular signature object

```{r}
str(reg_result)
```

Typical objects contain:

- `signature`
- `plots`
- `params`
- `metadata`
- `call`

---

### Survival analysis object

```{r}
str(surv_result)
```

Typical objects contain:

- `signature`
- `score_cutoff`
- `expression_data`
- `clinical_data`
- `survival`
- `roc_result`
- `ph_result`
- `plots`
- `params`
- `metadata`
- `call`

---

### Complete workflow object

```{r}
str(complete_result)
```

Typical objects contain:

- `regression`
- `survival`
- `params`
- `metadata`
- `call`

---

## Logging

Logging can be enabled by setting `log = TRUE`:

```{r}
rebootRegression(..., log = TRUE)
rebootSurvival(..., log = TRUE)
rebootComplete(..., log = TRUE)
```

Minimalist `.log` files containing timestamps and execution steps are automatically written to the working directory using module-specific filenames.

---

## Running a full example script

A complete example script is also available within the package:

```{r}
script_path <- system.file("scripts", "run_example.R", package = "Reboot")

script_path
```

This script briefly demonstrates how to run the complete REBOOT workflow using example toy data.

---

## Reproducibility

REBOOT can save:

- Function calls
- Execution parameters
- Package version
- Timestamps
- Optional command-line invocation metadata

These metadata are stored within returned objects and exported JSON metadata files, as exemplified before using `write_reboot_metadata()`.

---

## Comprehensive summary report

REBOOT can generate a summary report containing:

- Metadata, such as function calls and parameters used
- Basic statistics regarding the input datasets
- Summary tables and figures generated throughout the pipeline

The aim of the report is to help users organize and interpret the analysis, as exemplified before using `write_reboot_report()`.

---

## Saving and loading REBOOT objects (RDS)

REBOOT provides utilities to save and reload complete analysis objects in RDS format. This allows users to:

- Persist long-running bootstrap analyses
- Resume work in later R sessions
- Share complete REBOOT results between environments
- Ensure full reproducibility of analytical workflows

### Saving REBOOT objects

The function `write_reboot_rds()` can be used to save any REBOOT object
(`reboot_signature`, `reboot_survival`, or `reboot_complete`) into a compressed RDS file.

```{r}
write_reboot_rds(complete_result, "my_complete_result.rds")
```

By default, compression is set to "xz", ensuring compact file sizes suitable for large genomic analyses.

---

### Loading REBOOT objects

Previously saved REBOOT objects can be restored using the function `read_reboot_rds()`:

```{r}
loaded_object <- read_reboot_rds("my_complete_result.rds")

summary(loaded_object)
```

The function automatically validates that the loaded file contains a valid REBOOT object and preserves its S3 class:

- `reboot_signature`
- `reboot_survival`
- `reboot_complete`

This ensures full compatibility with all downstream methods (print, summary, coef, plot, etc.).

---

## Notes

- Runtime depends strongly on the number of bootstrap iterations and dataset size.
- Large values of `bootstrap` (e.g., 10,000) may substantially increase runtime.
- For exploratory analyses, smaller datasets and bootstrap values are recommended.
- The package is primarily designed for high-dimensional transcriptomic survival analyses.
- Users are encouraged to further investigate the REBOOT parameters and the generated signature.

---

## Main dependencies

Core package dependencies include:

- survival
- survivalROC
- survminer
- penalized
- ggplot2
- OptimalCutpoints
- sjstats

See the DESCRIPTION file for a complete and updated list of dependencies.

---

## See also

- GitHub repository: https://github.com/galantelab/reboot
- Documentation website: https://galantelab.github.io/reboot/

---

## Session information

```{r}
sessionInfo()
```
